---
title: "Homework 2"
author: "Eric Brewster"
date: "July 11, 2024"
output:
  pdf_document:
    extra_dependencies: amsmath
  html_document:
    df_print: paged
---
 
```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(echo = TRUE, eval=TRUE, tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

# Problem 1
Assuming, Y1,...,Yn, are i.i.d. N(mu, sigma) (normally distributed), construct a 100(1 - alpha)% upper CL for sigma squared (population variance).

This problem requires the existence of a pivotal quantity. Given the normal distribution and the unknown mean mu/variance sigma squared, by Theorem 7.3,
$$
\frac{\sum_{i=1}^n (Y_i - \bar{Y})^2}{\sigma^2} = \frac{(n - 1) S^2}{\sigma^2}
$$

the above expression has a chi-square distribution with n minus 1 degrees of freedom. The next step is to use the pivotal method to determine the upper confidence level.

$$
P\left( \chi^2_L \leq \frac{(n - 1)S^2}{\sigma^2} \leq \chi^2_U \right) = 1 - \alpha
$$

The chi-square distribution is not symmetric, so choosing lower and upper values can be difficult. The compromise is to choose tails of equal area; as a result, the upper confidence level is the alpha valued halved.

$$
P\left( \chi^2_{1-(\alpha/2)} \leq \frac{(n - 1)S^2}{\sigma^2} \leq \chi^2_{\alpha/2} \right) = 1 - \alpha
$$

When you take the reciprocal of each term, you also switch the signs; in this case, this means the two tails switch places. Multiplying each term by the new denominator of the middle term (n-1 /S^2), the middle term becomes sigma squared, the value asked for in the problem.

$$
P\left( \frac{(n - 1)S^2}{\chi^2_{\alpha/2}} \leq \sigma^2 \leq \frac{(n - 1)S^2}{\chi^2_{1-\alpha/2}} \right) = 1 - \alpha
$$


The upper confidence limit for population variance is
$$
\frac{(n - 1)S^2}{\chi^2_{1 - (\alpha/2)}}
$$


# Problem 2: 8.39
Suppose that the random variable Y has a gamma distribution with parameters alpha = 2 and an
unknown beta. In Exercise 6.46, you used the method of moment-generating functions to prove a
general result implying that 2Y/beta has a chi-square distribution with 4 degrees of freedom (df). Using 2Y/beta as a pivotal quantity, derive a 90% confidence interval for beta.

Using the given pivotal quantity, this is the interval to determine the lower and upper values.
$$
P(a \leq 2 \frac Y \beta \leq b) = 0.9
$$

To more quickly solve the algebra and calculus used in the textbook, R can determine the bounds using qchisq() for both lower and upper. 
$$
P(2 \frac Y \beta < a) = P(2 \frac Y \beta > b) = 0.05
$$

```{r}
qchisq(0.95, 4, lower.tail = FALSE)
```
```{r}
qchisq(0.05, 4, lower.tail = FALSE)
```
Substituting in the values of a and b gives the probability equation: 
$$
P(0.71 \leq 2 \frac Y \beta \leq 9.49) = 0.9
$$

Isolating beta through algebra gives the desired confidence interval. To do this, 

$$
P(0.71 \leq 2 \frac Y \beta \leq 9.49) = 0.9
$$
$$
P(0.71\beta \leq 2Y \leq 9.49\beta) = 0.9
$$

Separating the expression into two inequalities and isolating the beta, 
$$
\beta \leq \frac{2Y}{0.71}, \beta \geq \frac{2Y}{9.49}
$$

Given that 2Y/0.71 will be greater than 2Y/9.49, merge into one inequality again,

$$
P(\frac{2Y}{9.49} \leq \beta \leq \frac{2Y}{0.71}) = 0.9
$$

The 90% confidence interval for beta is
$$
(\frac{2Y}{9.49}, \frac{2Y}{0.71})
$$
These limits will include the true (unknown) value of beta for 90% of the values from repeated sampling from this chi-square distribution.


# Problem 3: 8.47
Refer to Exercise 8.46. Assume that Y1, Y2, . . . , Yn is a sample of size n from an exponential
distribution with mean theta.

## Part B
Use the pivotal quantity 
$$
2\sum_{i=1}^n \frac{Y_i}{\theta}
$$
to derive a 95% confidence interval for theta.

The problem asks for the interval ends a and b in
$$
P(a \leq 2\sum_{i=1}^n \frac{Y_i}{\theta} \leq b) = 0.95
$$

To rephrase, this means finding a and b so there are two tails each with 0.025% as shown below:
$$
P\left(2 \sum_{i=1}^{n} \frac {Y_i} \theta < a\right),
P\left(2 \sum_{i=1}^{n} \frac {Y_i} \theta > b\right)
$$

If the degrees of freedom were known, the chi-square table could be used (or use qchisq() in R). As the degrees of freedom are not (exactly) known, substitute in the 2n from the problem definition into a chi-square variable. For a and b, this translates to
$$
a = \chi^2_{2n} (0.975), b = \chi^2_{2n} (0.025)
$$

Returning to the problem at hand, the solution is to isolate theta in the central term. The first step inserts the chi-square equivalent of a/b and brings the constant of 1 divided by theta before the summation.
$$
P(\chi^2_{2n} (0.975) \leq \frac 2 \theta \sum_{i=1}^n {Y_i} \leq \chi^2_{2n} (0.025)) = 0.95
$$

Next, multiply each term by theta:
$$
P(\theta\chi^2_{2n} (0.975) \leq  2 \sum_{i=1}^n {Y_i} \leq \theta \chi^2_{2n} (0.025)) = 0.95
$$

Taking reciprocals of the terms means switching the direction of the inequality. In this case, I switched left and right terms.
$$
P(\frac{2 \sum_{i=1}^n {Y_i}}{\chi^2_{2n} (0.025)} \leq \theta \leq \frac {2 \sum_{i=1}^n {Y_i}}{\chi^2_{2n} (0.975)}) = 0.95
$$

The 95% confidence interval,
$$
(\frac{2 \sum_{i=1}^n {Y_i}}{\chi^2_{2n} (0.025)}, \frac {2 \sum_{i=1}^n {Y_i}}{\chi^2_{2n} (0.975)})
$$
will contain the true (unknown) values of theta for 95% of the values obtained from Y by sampling from the exponential distribution.


## Part C
If a sample of size n = 7 yields y = 4.77, use the result from part (b) to give a 95%
confidence interval for theta.

Write out the result from part b, recalling that the sum of samples represents y-bar, or the sample mean, multiplied by the number of samples. As the problem gives the sample mean and sample number, substitute the value in.
$$
(\frac{2n \bar Y} {\chi^2_{2n} (0.025)}, \frac {2n \bar Y}{\chi^2_{2n} (0.975)})
= (\frac{2(7)(4.77)} {\chi^2_{14} (0.025)}, \frac {2(7)(4.77)}{\chi^2_{14} (0.975)})
$$
Using the same code as before, input degrees of freedom and desired tail for the chi-squared value.
```{r}
qchisq(0.975, 14, lower.tail = FALSE)
```

```{r}
qchisq(0.025, 14, lower.tail = FALSE)
```
Substituting in the chi-square values,
$$
(\frac{2(7)(4.77)} {26.12}, \frac {2(7)(4.77)}{5.63})
= (2.56, 11.86)
$$

gives the 95% confidence interval for theta of (2.56, 11.86).

# Problem 4: 8.56
Is America’s romance with movies on the wane? In a Gallup Poll of n = 800 randomly chosen
adults, 45% indicated that movies were getting better whereas 43% indicated that movies were
getting worse.

## Part A
Find a 98% confidence interval for p, the overall proportion of adults who say that movies
are getting better.

Below is the formula for construsting a confidence interval based on successes in a binomial experiment:
$$
\hat{p} \pm z_{\alpha/2} \sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}
$$
Determining the z-value using the t-table, given the binomial distribution:
$$
z_{.01} = 2.326, z_{.99} = 2.326
$$

Substitute in the values given in the problem definition. P-hat represents the proportion of successes, in the case the proportion who say movies are getting better and n being the number of respondents:
$$
0.45 \pm 2.33 \sqrt{\frac{0.45(1 - 0.45)}{800}}
= (0.41, 0.49)
$$

The 98% confidence interval for p is (0.41, 0.49). This represents the proportion of adults who say that movies are getting better in 98% of iterations of using this procedure.

## Part B
Does the interval include the value p = .50? Do you think that a majority of adults say
that movies are getting better?

The interval does not include the value p= .50, so I do not think a majority of adults say that movies are getting better. The procedure that generates the interval does not include 0.50 in approximately 98% of instances where this procedure is used, so there is not reason to believe that majority of adults say movies are getting better.


# Problem 5: 8.85
Two new drugs were given to patients with hypertension. The first drug lowered the blood
pressure of 16 patients an average of 11 points, with a standard deviation of 6 points. The
second drug lowered the blood pressure of 20 other patients an average of 12 points, with a
standard deviation of 8 points. Determine a 95% confidence interval for the difference in the
mean reductions in blood pressure, assuming that the measurements are normally distributed
with equal variances.

The equation below is used find the confidence interval for estimating the true mean difference between two populations:
$$
(\bar y_1 - \bar y_2) \pm t_{\alpha/2} s_p \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}
$$

Filling in known terms from the problem definition:
$$
(11 - 12) \pm t_{\alpha/2} s_p \sqrt{\frac{1}{16} + \frac{1}{20}}
$$

Using the t-table:
$$
t_{0.025, 34} \approx 2.03 
$$

Determining the pooled estimator for variance:
$$
S_p^2 = \frac{(n_1 - 1)S_1^2 + (n_2 - 1)S_2^2}{n_1 + n_2 - 2}
= \frac{(16 - 1)6^2 + (20 - 1)8^2}{16 + 20 - 2}
= 51.647, 
s_p = 7.19
$$

Putting all the variables together:
$$
(-1) \pm (2.03)(7.19)(\frac{1}{6})
= (-5.89, 3.89)
$$

The final interval is (-5.89, 3.89). This is the interval in which we are 95% certain that the true mean difference between blood pressure reduction by medicine lies. Because the interval contains both positive and negative values, neither drug can be said to produce a mean blood pressure reduction that differs from the other.

