---
title: "In-Class Project 3"
author: "Eric Brewster"
date: "3/1/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Part 1 - Response Surface to optimize (maximize) guanadine hydrochloride (GuHcl) extraction in production of murals.
# Y = GuHcl extraction efficiency (%)  extEff_Y
# X1 = GuHcl concentation (M)   GuHCl
# X2 = Extraction Time (minutes)  extTime
# X3 = Extraction Temperature (C)   extTemp
# X4 = Ultrasonic Power (W)   ultraPwr

# Read in data and display "head" and "tail"

```{r part1a}
murals <- read.csv("https://www.stat.ufl.edu/~winner/data/murals_rsm.csv")
attach(murals); head(murals); tail(murals)
```

# Plot Y versus each factor 1-at-a-time

```{r part1b}
par(mfrow=c(2,2))
plot(extEff_Y ~ GuHCl, pch=16)
plot(extEff_Y ~ extTime, pch=16)
plot(extEff_Y ~ extTemp, pch=16)
plot(extEff_Y ~ ultraPwr, pch=16)
par(mfrow=c(1,1))
```

# Fit the second order response surface

```{r part1cde}
library(rsm)
mod1 <- rsm(extEff_Y ~ SO(GuHCl, extTime, extTemp, ultraPwr))
summary(mod1)
```
d) Consider t-tests for individual parameters. Are any not significant at 0.10 level? If so, which ones?
All four tests for individual parameters are significant at the 0.10 level (significant at even lower levels as well).

e) Consider Sequential F-tests for groups of predictors (Linear, 2-Factor Interactions, Quadratics). Are any not significant at 0.05 level? If so, which ones?
GuHCl:extTemp and GuHCl:ultraPwr not significant at the 0.05 level.


# Fit Reduced Model and compare with Full

```{r part1fg}
mod2 <- rsm(extEff_Y ~ FO(GuHCl,extTime,extTemp,ultraPwr) + TWI(extTime, extTemp,ultraPwr) + TWI(GuHCl,extTime) + PQ(GuHCl,extTime,extTemp,ultraPwr))
summary(mod2)
anova(mod2,mod1)
```
g) Test whether the Reduced Model fits as well as the Full Model
Considering that the full model's lack-of-fit p-value is 0.105157 and the reduced model's lack-of-fit p-value is 0.090585, neither is significant at the 0.05 level. There is not strong evidence that the full model is any better fit than reduced.


# 3-D Surface Plots

```{r part1h}
par(mfrow=c(2,3))
persp(mod2, ~ GuHCl+extTime+extTemp+ultraPwr, col="green",
   at=summary(mod2)$canonical$xs)
par(mfrow=c(1,1))
```

# Contour Plots
```{r part1i}
par(mfrow=c(2,3))
contour(mod2, ~ GuHCl+extTime+extTemp+ultraPwr, 
   at=summary(mod2)$canonical$xs, image=TRUE)
par(mfrow=c(1,1))
```


# Detach dataframe

```{r}
detach(murals)
```

### Part 2 - Luminance of Egyptian Cotton varieties by grade

# Read in data

```{r part2ab}
ec <- read.csv("http://users.stat.ufl.edu/~winner/data/egyptcttn.csv")
attach(ec)
head(ec); tail(ec)

variety1 <- rep(2:6,each=4)   # G69=2, G67=3, G70=4, G68=5, M=6
```

# Plot luminance vs lnGrade with different colors for varieties

```{r part2c}
plot(luminance ~ lnGrade, col=variety1, pch=16)
```

# Fit simple linear regression and plot

```{r part2d}
(TSS <- sum((luminance-mean(luminance))^2))
ec.mod1 <- lm(luminance ~ lnGrade)
summary(ec.mod1)
anova(ec.mod1)
confint(ec.mod1)

coef(ec.mod1)

X.grid <- seq(0,5,0.1)
yhat1 <- coef(ec.mod1)[1] + coef(ec.mod1)[2]*X.grid
plot(luminance ~ lnGrade, pch=16, xlim=c(0,5),  ylim=c(82,92), col=variety1)
lines(X.grid,yhat1, col="blue", lwd=2)
```

# Fit additive model (Different intercepts, common slopes) and plot


```{r part2e}
ec.mod2 <- lm(luminance ~ lnGrade + giza69 + giza67 + giza70 + giza68)
summary(ec.mod2)
anova(ec.mod2)
confint(ec.mod2)

coef(ec.mod2)
X.grid <- seq(0,5,0.1)
yhat1 <- coef(ec.mod2)[1] + coef(ec.mod2)[2]*X.grid
yhat2 <- yhat1 + coef(ec.mod2)[3]
yhat3 <- yhat1 + coef(ec.mod2)[4]
yhat4 <- yhat1 + coef(ec.mod2)[5]
yhat5 <- yhat1 + coef(ec.mod2)[6]

plot(luminance ~ lnGrade, pch=16, col=variety1)
lines(X.grid,yhat1, col=6, lwd=2)
lines(X.grid,yhat2, col=2, lwd=2)
lines(X.grid,yhat3, col=3, lwd=2)
lines(X.grid,yhat4, col=4, lwd=2)
lines(X.grid,yhat5, col=5, lwd=2)

```

# Fit interaction model (Different intercepts and slopes) and plot

```{r part2f}
ec.mod3 <- lm(luminance ~ lnGrade + giza69 + giza67 + giza70 + giza68 +
     lnGrG69 + lnGrG67 + lnGrG70 + lnGrG68)
summary(ec.mod3)
anova(ec.mod3)
confint(ec.mod3)

X.grid <- seq(0,5,0.1)
yhat1 <- coef(ec.mod3)[1] + coef(ec.mod3)[2]*X.grid
yhat2 <- yhat1 + coef(ec.mod3)[3] + coef(ec.mod3)[7]*X.grid
yhat3 <- yhat1 + coef(ec.mod3)[4] + coef(ec.mod3)[8]*X.grid
yhat4 <- yhat1 + coef(ec.mod3)[5] + coef(ec.mod3)[9]*X.grid
yhat5 <- yhat1 + coef(ec.mod3)[6] + coef(ec.mod3)[10]*X.grid

plot(luminance ~ lnGrade, pch=16, col=variety1)
lines(X.grid,yhat1, col=6, lwd=2)
lines(X.grid,yhat2, col=2, lwd=2)
lines(X.grid,yhat3, col=3, lwd=2)
lines(X.grid,yhat4, col=4, lwd=2)
lines(X.grid,yhat5, col=5, lwd=2)
```

 
# Test between the various models

```{r part2gh}
anova(ec.mod1,ec.mod2) # simple and additive
anova(ec.mod2,ec.mod3) # additive and interaction
anova(ec.mod1,ec.mod3) # simple and interaction

```
g and h) Test between the models in e) and d) for a variety effect (allowing different intercepts, controlling for lnGrade), Test between the models in f) and e) for an interaction effect (slopes differ among varieties)

Test between e (additive multiple linear) and d (simple linear) for variety effect: 1.066e-07 p-value, significant at the 0.05 value.

Test between f (interaction multiple linear) and e (additive multiple linear) for interaction effect: 0.02313 p-value, significant at 0.05 but not 0.01. Not as significant as variety effect.

Also tested: f (interaction multiple linear) and d (simple linear): 7.114e-07 p-value, significant at the 0.05 value.


# Fit intermediate model (Combine M&G67 and G67&G70, making G68 the reference category)

```{r part2i}
mg69 <- ifelse(variety=="Menoufi" | variety=="Giza69",1,0)
g6770 <- ifelse(variety=="Giza67" | variety=="Giza70",1,0)
ec.mod4 <- lm(luminance ~ lnGrade + mg69 + g6770 + I(lnGrade*mg69) + I(lnGrade*g6770))
summary(ec.mod4)

anova(ec.mod4, ec.mod3)

yhat1 <- coef(ec.mod4)[1] + coef(ec.mod4)[2]*X.grid
yhat2 <- yhat1 + coef(ec.mod4)[4] + coef(ec.mod4)[5]*X.grid
yhat3 <- yhat1 + coef(ec.mod4)[4] + coef(ec.mod4)[6]*X.grid

variety1a <- ifelse(variety1 ==2 | variety1==6, 2, ifelse(variety1 <= 4,3,4))
variety1a
plot(luminance ~ lnGrade, pch=16, col=variety1a)
lines(X.grid,yhat1, col=4, lwd=2)
lines(X.grid,yhat2, col=2, lwd=2)
lines(X.grid,yhat3, col=3, lwd=2)
```


i) If appropriate, fit an intermediate model, plot it, and test whether it is adequate
When testing between intermediate model and mod3 (interaction multiple linear), the p-value is 0.0245. Significant at 0.05 but not 0.01
