---
title: "HW1"
author: "Eric Brewster"
date: "2024-01-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r AUC setup}
spdata <- read.csv("http://www.stat.ufl.edu/~winner/data/bioequiv_sulf.csv")
attach(spdata); names(spdata)
spdata
spdata.AS <- spdata[measure==2 & drug==1,]
detach(spdata)
attach(spdata.AS)

## Create a variable that is AUC for the drug sulfadoxine and a 
##   variable for formulation

AUC.sulf <- y[measure==2 & drug==1]    # measure=AUC, drug=sulfadoxine
form.AUC.sulf <- form[measure==2 & drug==1] # form=1 if Test, form=2 if Ref
form.AUC.sulf <- factor(form.AUC.sulf, levels=1:2, labels=c("T","R"))
cbind(form.AUC.sulf, AUC.sulf)

## Side-by-side Boxplots
plot(AUC.sulf ~ form.AUC.sulf, main="Sulfadoxine AUC by Formulation")
```

```{r 1a}
AUCtTest <- t.test(AUC.sulf[form.AUC.sulf == "T"], AUC.sulf[form.AUC.sulf == "R"])
AUCtTest
```


```{r 1b}
AUCvarTest <- var.test(AUC.sulf ~ form.AUC.sulf)
AUCvarTest
```


```{r 1c}
AUCmodel <- lm(AUC.sulf ~ form.AUC.sulf)
summary(AUCmodel)
```

```{r Q3 setup}
latlongData <- read.csv("C:\\Users\\14048\\Downloads\\spain_latlong.csv")
library(ggplot2)
```


```{r 3a}
ggplot(latlongData, aes(x = Lat_ptol, y = Lat_wgs)) +
  geom_point() +
  labs(title = "Scatterplot of GIS versus Ptolemy for Latitude",
       x = "Ptolemy's Latitude Assessment",
       y = "GIS Latitude Assessment")
```


```{r 3b}
lm_model_lat <- lm(Lat_wgs ~ Lat_ptol, data = latlongData)
```


```{r 3c}
summary(lm_model_lat)
```
T-Test
Null H0: No association between GIS and Ptolemy for latitude
Alternative H1: Positive association between the two
Test stat: 20.904
p-value: < 2e-16
Conclusion: The very miniscule p-value indicates that we should reject the null hypothesis and conclude there is a positive correlation between the GIS and Ptolemy latitudes


```{r 3d}
confint(lm_model_lat, "Lat_ptol")
```


```{r 3e}
anova(lm_model_lat)
```
F-Test
Null H0: No relationship between predictor and response
Alternative H1: At least one coefficient is not zero, so there is a relationship
Test stat: 436.96
p-value: < 2.2e-16
Conclusion: The very low p-value, so reject the null hypothesis and conclude that there is a significant relationship between predictor and response


```{r 3f}
cor(latlongData$Lat_ptol, latlongData$Lat_wgs)
cor(latlongData$Lat_ptol, latlongData$Lat_wgs)^2
```

