---
title: "In-Class Project 5"
author: "Eric Brewster"
date: "2024-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Read in City Annual Temp Data
### Sample 1 city to analyze

```{r readdata}

city.data <- read.csv("https://www.stat.ufl.edu/~winner/data/allcities_annual_19702020.csv")
head(city.data); tail(city.data)

set.seed(0542)

(num.city <- length(unique(city.data$city_ID)))

(city.sample <- sample(num.city, 1))

cities <- unique(city.data$City)
(my.city <- cities[city.sample]) # Cheyenne

city.temp <- city.data[city.data$city_ID==city.sample,]
head(city.temp); tail(city.temp)
```

### OLS Estimates, t-tests, CI's for Your City

```{r OLS}

mod.OLS <- lm(aveTemp ~ I(year-1970), data=city.temp)
summary(mod.OLS)
confint(mod.OLS)
b0_OLS <- summary(mod.OLS)$coef[1,1]
b1_OLS <- summary(mod.OLS)$coef[2,1]
s.b1_OLS <- summary(mod.OLS)$coef[2,2]
```
$\hat{Y}$ = `r b0_OLS` + `r b1_OLS`X   $s\{b1_{OLS}\}$ = `r s.b1_OLS`


### Durbin-Watson Test and Permutation Based P-value for AR(1) Model
$$ \epsilon_t = \rho \epsilon_{t-1} + u_t \quad \{u_t\} \bot \{\epsilon_{t-k}\} \quad k>0  \qquad \qquad   H_0:\rho=0 \qquad H_A:\rho \neq 0$$

```{r DW}

e.OLS <- resid(mod.OLS)
n <- length(e.OLS)
DW.OLS <- sum((e.OLS[2:n]-e.OLS[1:(n-1)])^2) / sum(e.OLS^2)

n.perm <- 9999
set.seed(32611)
DW.OLS
DW.OLS.perm <- rep(0, n.perm)
SSE <- sum(e.OLS^2)
for (i1 in 1:n.perm) {
  perm <- sample(n, n, replace=F)
  e.OLS.perm <- e.OLS[perm]
  DW.OLS.perm[i1] <- sum((e.OLS.perm[2:n]-e.OLS.perm[1:(n-1)])^2)/SSE
}

p.DW.OLS <- (sum(DW.OLS.perm <= DW.OLS | DW.OLS.perm >= 4-DW.OLS)+1) / (n.perm+1)
p.DW.OLS


summary(DW.OLS.perm)
hist(DW.OLS.perm, breaks=seq(min(DW.OLS.perm)-.001,max(DW.OLS.perm)+.001,length=40))
abline(v=DW.OLS, col="blue", lwd=2)
abline(v=4-DW.OLS, col="blue", lwd=2)

```
Test Statistic: 1.775379     P-value: 0.414        Conclusion: Not enough evidence to reject the null hypothesis of no autocorrelation 

### Fit Estimated Generalized Least Squares based on Yule Walker equations

```{r EGLS_YW}

X <- model.matrix(mod.OLS)
X[1:12,]
Y <- city.temp$aveTemp
p <- ncol(X)

(gamma0 <- sum(e.OLS^2) / n)
(gamma1 <- sum(e.OLS[2:n]*e.OLS[1:(n-1)])/n)
(rho <- gamma1 / gamma0)
(sigma2 <- gamma0 - rho*gamma1)

T.inv <- diag(n)
T.inv[1,1] <- sqrt(1-rho^2)
for(i1 in 2:n)  T.inv[i1,(i1-1)] <- -rho

X.s <- T.inv %*% X
Y.s <- T.inv %*% Y

beta.egls <- solve(t(X.s) %*% X.s) %*% t(X.s) %*% Y.s

e.egls <- as.vector(Y.s - X.s %*% beta.egls)
(DW.egls <- sum((e.egls[2:n]-e.egls[1:(n-1)])^2) / sum(e.egls^2))

(sigma2_e <- (t(Y-X%*%beta.egls) %*% t(T.inv) %*% T.inv %*% 
   (Y-X%*%beta.egls)) / (n-p-1))
(sum(e.egls^2) / (n-p-1))

V.beta.egls <- sigma2_e[1,1] * solve(t(X.s) %*% X.s)
SE.beta.egls <- sqrt(diag(V.beta.egls))

s2 <- (gamma0-rho*gamma1)/(n-p-1)
SE.rho <- sqrt(s2/gamma0)

rho.out <- cbind(rho, SE.rho, rho/SE.rho, 2*(1-pt(abs(rho/SE.rho),n-p-1)))
colnames(rho.out) <- c("Estimate", "Std Err", "t", "P(>|t|)")
rownames(rho.out) <- c("rho")
round(rho.out, 4)

beta.out <- cbind(beta.egls,SE.beta.egls,beta.egls/SE.beta.egls,
  2*(1-pt(abs(beta.egls/SE.beta.egls),n-p-1)))
colnames(beta.out) <- c("Estimate", "Std Err", "t", "P(>|t|)")
rownames(beta.out) <- c("Intercept", "Series Month")
round(beta.out, 4)

win.graph(height=5.5, width=7.0)
plot(e.egls, type="o", col="blue", 
   main="Residuals from Linear Trend Model (EGLS)")
abline(h=0, col="red", lwd=2)
```

$$ \hat{\rho}= 0.1087    \qquad  \qquad         b_1 = 0.0472  \qquad      \qquad   s\{b_1\} = 0.0123  $$

### Fit Estimated Generalized Least Squares based on Maximum Likelihood and compare with OLS

```{r EGLS_ML}

library(nlme)

mod.EGLS <- gls(aveTemp ~ I(year-1970), method="ML", correlation=corARMA(p=1), data=city.temp)
mod.OLS1 <- gls(aveTemp ~ I(year-1970), method="ML", data=city.temp)

summary(mod.EGLS)
anova(mod.OLS1, mod.EGLS)

```

Based on Likelihood Ratio Test for $H_0:\rho=0$ 
Test Statistic: 0.5991064      P-value: 0.4389   Conclusion: Not enough evidence to reject the null hypothesis that rho = 0     