---
title: "Homework2STA4210"
author: "Eric Brewster"
date: "2024-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r start}
hw2 <- read.csv("https://www.stat.ufl.edu/~winner/data/explosives1.csv")
attach(hw2); names(hw2)
group.BF <- ifelse(coupling <= 4, 1, 2)
group.BF
```


```{r a}
fit1 <- lm(defGalv ~ areaWire)
summary(fit1)
resid(fit1)
```
yhat = b0 + b1x = 184.43569 - 0.69537x



```{r b}
plot(areaWire, defGalv, main = "Y vs X", xlab = "Area of Wires", ylab = "Deflection of Galvonometer")
abline(fit1, col = "red")
```


```{r c}
plot(areaWire, resid(fit1), main = "Residuals vs X", xlab = "Area of Wires", ylab = "Residuals", col = "blue")
abline(h = 0, col = "red")
```


```{r d}
summary(fit1)
```
H0: Deflections are not associated with wire area (area wire coefficient = 0, B1 = 0)
HA: Wire area has a significant impact on deflections (B1 /= 0)
Test Statistic: -20.56
Rejection Region: t < -2.086 and t > 2.086
P-value: 6.39e-15


```{r e}
library(car)
qqPlot(resid(fit1), main = "Normal Probability Plot of Residuals")
```


```{r f}
shapiro.test(resid(fit1))
```
H0: The residuals are normally distributed
p-value: 0.09004


```{r g}
res1 <- resid(fit1)
n1 <- length(res1[group.BF == 1])
n2 <- length(res1[group.BF == 2])

median_e1 <- median(res1[group.BF == 1])
median_e2 <- median(res1[group.BF == 2])

median_e <- c(rep(median_e1,n1),rep(median_e2,n2))
median_e

d.BF <- abs(res1 - median_e)
cbind(group.BF, d.BF)

t.test(d.BF ~ group.BF, var.equal=TRUE)

dbar1 <- mean(d.BF[group.BF == 1])
dbar2 <- mean(d.BF[group.BF == 2])

var1 <- var(d.BF[group.BF == 1])
var2 <- var(d.BF[group.BF == 2])

var.p <- ((n1-1)*var1 + (n2-1)*var2) / (n1 + n2 -2)

t.BF <- (dbar1 - dbar2) / sqrt(var.p*(1/n1+1/n2))
p.t.BF <- 2*(1-pt(abs(t.BF), n1+n2-2))

BF.out  <- cbind(n1+n2-2, dbar1-dbar2, t.BF, p.t.BF)
colnames(BF.out) <- c("df", "Mean Diff", "t*", "2P(>|t*|)")
rownames(BF.out) <- c("BF Test")
round(BF.out,4)

library(car)
leveneTest(res1, factor(group.BF), "median")

BF.aov <- aov(d.BF ~ factor(group.BF))
summary(BF.aov)

anova(lm(d.BF ~ factor(group.BF)))

res.df <- data.frame(res1, group.BF)


########## This does NOT perform the test in the same manner!!!!!!
library(onewaytests)
bf.test(res1 ~ factor(group.BF), data=res.df)

```
H0: The variances of the groups are equal
HA: The variances of the groups are not equal
Test Statistic: 0.5052253
Rejection Region: test-stat > 6.042
P-value: 0.4865665


```{r h}
library(lmtest)
bptest(fit1)
```
H0: There is constant variance in the model
HA: There exists non-constant variance in the model 
Test Statistic: 2.5908
Rejection Region: test-stat > 5.0239
P-value: 0.1075


```{r i}
anova(fit1, test = "F")
```

H0: No lack of fit in the full model
HA: There is a lack of fit in the full model (not significantly more predictive than restricted model)
Test Statistic: 422.6
Rejection Region: F > 5.8715
P-value: 6.386e-15


```{r j}
durbinWatsonTest(fit1)
```
H0: rho = 0
HA: rho /= 0
Test Statistic: 0.8957278
P-value: 0


```{r k}
library(MASS)
bc <- boxcox(fit1)
(lambda <- bc$x[which.max(bc$y)])
```
lambda-hat: -0.222
95% CI shown in plot


```{r l}
confint(fit1, level = 0.95)
```
B0 Interval: (178.3545618, 190.5168123)
B1 Interval: (-0.7659248, -0.6248066)


```{r m}
predict(fit1, newdata = data.frame(areaWire = 115), interval = "confidence", level = 0.95)
```
Interval: (100.0478, 108.8895)

```{r n}
(X <- as.matrix(fit1$model$areaWire))
(Y <- as.matrix(fit1$model$defGalv))
(XtX <- t(X) %*% X)
(XtY <- t(X) %*% Y)
(inv_XtX <- solve(XtX))
(b <- inv_XtX %*% XtY)
(MSE <- deviance(fit1)/df.residual(fit1))
(s2_b <- vcov(fit1))
```

