---
title: "In-Class Assignment 4"
author: "Eric Brewster"
date: "4/5/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Part 1 - NFL Combine 2014 Data

## Part 1a Read in Data, Fit OLS Regression, Obtain Residual Plots and Breusch-Pagan Test
```{r part1a}
nflcomb <- read.csv("http://www.stat.ufl.edu/~winner/sta4210/mydata/nfl_combine.csv",
       header=TRUE)
attach(nflcomb); names(nflcomb)

nfl.mod1 <- lm(Weight ~ Height + ArmLng + HandLng)
summary(nfl.mod1)

par(mfrow=c(2,2))
plot(resid(nfl.mod1) ~ predict(nfl.mod1), pch=16)
abline(h=0, col="red", lwd=2)
plot(resid(nfl.mod1) ~ Height, pch=16)
abline(h=0, col="red", lwd=2)
plot(resid(nfl.mod1) ~ ArmLng, pch=16)
abline(h=0, col="red", lwd=2)
plot(resid(nfl.mod1) ~ HandLng, pch=16)
abline(h=0, col="red", lwd=2)
par(mfrow=c(1,1))

library(lmtest)
bptest(nfl.mod1, studentize=FALSE)
```


## Part 1b - Estimated iteratively reweighted least squares (IRLS) with:

$$ \sigma_i^2 = \sigma^2\mu_i^{2\delta} \qquad \Longrightarrow \qquad \log\left(\sigma_i^2\right)=\log\left(\sigma^2\right)+\delta\log\left(\mu_i^2\right)$$
```{r part1b}
#### Matrix form (using lm for log(e^2) , log(y-hat^2) regression) ###########

n <- length(Weight)
X0 <- rep(1,n)
X <- as.matrix(cbind(X0,Height,ArmLng,HandLng))
Y <- as.matrix(Weight)
p <- ncol(X)

#### Fit original regression, and regress functions of |e| on Y-hat

b.ols <- solve(t(X) %*% X) %*% t(X) %*% Y           # b=(X'X)^(-1)X'Y
mse.o <- (t(Y-X%*%b.ols) %*% (Y-X%*%b.ols))/(n-p)   # MSE=(Y-Xb)"(Y-Xb)/(n-p)
s2.b.ols <- mse.o[1,1]*solve(t(X) %*% X)            # s2{b}=MSE*(X'X)^(-1)
s.b.ols <- sqrt(diag(s2.b.ols))                     # s{b}=sqrt(diag(b}}
t.b.ols <- b.ols/s.b.ols                            # t=b/s{b}
p.b.ols <- 2*(1-pt(abs(t.b.ols), n-p))              # P-value

ols.out <- cbind(b.ols, s.b.ols, t.b.ols, p.b.ols)
colnames(ols.out) <- c("b_ols", "Std. Error", "t*", "P(>|t*|)")
rownames(ols.out) <- c("Intercept", "Height", "Arm Length", "Hand Length")
round(ols.out,4)

#### Plot  log(e^2) vs log(Y-hat^2)
yhat.ols <- X %*% b.ols
e.ols <- Y - yhat.ols

plot(log(e.ols^2) ~ log(yhat.ols^2), pch=16)
abline(lm(log(e.ols^2) ~ log(yhat.ols^2)))

cor(log(e.ols^2) , log(yhat.ols^2))

# Fit regression of log(e^2) on log(Y-hat^2)
e.reg.ols <- lm(log(e.ols^2) ~ log(yhat.ols^2))
summary(e.reg.ols)
s2.ols <- exp(predict(e.reg.ols))         # Predicted s_i^2
w.ols <- 1/s2.ols                         # WLS weights= 1/s^2

## Begin iterations to obtain WLS estimator b_w
b.old <- b.ols                       # Start iterations with OLS estimator
wm.old <- as.matrix(diag(w.ols))     # W = diagonal matrix with w_i=1/s_i^2
b.diff <- 100                        # Set high starting difference 
num.iter <- 0                        # Counter for number of iterations

# Keep iterating until (b_new-b_old)'(b_new-b_old) < .00001
while (b.diff > 0.00001) {
   num.iter <- num.iter+1           # Increment number of iterations
   # b_new = (X'WX)^(-1)X'WY
   b.new <- solve(t(X) %*% wm.old %*% X) %*% t(X) %*% wm.old %*% Y
   yhat.new <- X %*% b.new        # Yhat_new = Xb_new
   e.new <- Y - yhat.new          # new e = Y-Yhat_new
   # Create new weight matrix from regression of log(e^2) on log(yhat_new^2)
   wm.new <- as.matrix(diag(1/exp(predict(lm(log(e.new^2)~log(yhat.new^2))))))
   b.diff <- sum((b.new-b.old)^2) # sum of squared differences of b_new,b_old
   b.old <- b.new                 # b.old is assigned b.new
   wm.old <- wm.new               # Old weight matrix is assigned new weight matrix
}

# End of loop
num.iter       # Number of iterations needed

# Apply wm.new to get b.new (probably not necessary)
b.new <- solve(t(X) %*% wm.new %*% X) %*% t(X) %*% wm.new %*% Y
b.wls <- b.new          # Obtain b.wls as result from iterative process
wm.wls <- wm.new        # Obtain WLS matrix from iterative process

## MSE_wls = (Y-Xb-w)'W(Y-Xb_w)
mse.w <- (t(Y-X%*%b.wls) %*% wm.wls %*% (Y-X%*%b.wls))/(n-p)

# s2{b_w} = MSE*(X'WX)^(-1)
s2.b.wls <- mse.w[1,1]*solve(t(X) %*% wm.wls %*% X)
s.b.wls <- sqrt(diag(s2.b.wls))          # s{b_w} = sqrt(diag(s2{b_w}))
t.b.wls <- b.wls/s.b.wls                 # t = b_w/s{b}
p.b.wls <- 2*(1-pt(abs(t.b.wls), n-p))   # P-value

wls.out <- cbind(b.wls, s.b.wls, t.b.wls, p.b.wls)
colnames(wls.out) <- c("b_wls", "Std. Error", "t*", "P(>|t*|)")
rownames(wls.out) <- c("Intercept", "Height", "Arm Length", "Hand Length")

round(ols.out, 4)
round(wls.out, 4)
```


### Fit Maximum Likelihood Estimates w/ gls function in nlme package


```{r part1c}


## Using gls fuction in nlme package

library(nlme)
gls.mod.ols <- gls(Weight ~ Height+ArmLng+HandLng, method="ML")
summary(gls.mod.ols)

## Fit variance function model with s{e_i} = sigma*(mu_i^delta)
gls.mod.wls <- gls(Weight ~ Height+ArmLng+HandLng, method="ML",
    weights=varPower(form = ~fitted(.)))
summary(gls.mod.wls)

anova(gls.mod.ols,gls.mod.wls)

summary(predict(gls.mod.wls))
```

$$ \hat{\sigma}_i \quad = \quad \hat{\sigma}\hat{\mu}_i^{\hat{\delta}}\quad  = \quad .00055 \hat{\mu}_i^{1.98}   $$
###End of Part 1 and Begin of Part 2

Detach nflcomb and read in Bollywood box office data

```{r part2a}

detach(nflcomb)

bolly1 <- read.csv(
    "http://www.stat.ufl.edu/~winner/data/bollywood_boxoffice2.csv")
attach(bolly1); names(bolly1)
head(bolly1)
tail(bolly1)
```

## Part2b - Fit OLS model relating Gross Revenues to Budget and Plot Model and Residuals

```{r part2b}
Y1 <- Gross/100
X1 <- Budget/100

bolly.mod1 <- lm(Y1 ~ X1)
summary(bolly.mod1)
anova(bolly.mod1)

plot(Y1 ~ X1, pch=16, main="Gross vs Budget")
abline(bolly.mod1, col="blue", lwd=2)

plot(resid(bolly.mod1)~predict(bolly.mod1), pch=16, main="Residuals vs Predicted Values")
abline(h=0, col="red", lwd=2)

```

### Part 2c - Matrix form

```{r part2c}
n <- length(Y1)
X.mat <- cbind(rep(1, n), X1)
XpX <- t(X.mat) %*% X.mat
XpY <- t(X.mat) %*% Y1
betahat <- solve(XpX) %*% XpY

J_n <- matrix(rep(1/n,n^2), ncol=n)
I <- diag(n)
P <- X.mat %*% solve(XpX) %*% t(X.mat)

SSE <- t(Y1) %*% (I-P) %*% Y1
s2 <- SSE/(n-2)

V.betahat <- s2[1,1] * solve(XpX)
SE.betahat <- sqrt(diag(V.betahat))
t.betahat <- betahat/SE.betahat
p.betahat <- 2*(1-pt(abs(t.betahat),n-2))

ols.out <- cbind(betahat, SE.betahat, t.betahat, p.betahat)
colnames(ols.out) <- c("Estimate", "Std Err", "t", "P(>|t|)")
rownames(ols.out) <- c("Intercept", "Budget")
round(ols.out, 5)
```

## Part 2d - Robust (to heteroskedasticity) Standard Errors (Chapter 11, Slide 8)

```{r part2d}
library(lmtest)
bptest(bolly.mod1, studentize=FALSE)

E2 <- diag(resid(bolly.mod1)^2)

V.betahat.W <- solve(XpX) %*% t(X.mat) %*% E2 %*% X.mat %*% solve(XpX)
SE.betahat.W <- sqrt(diag(V.betahat.W))
t.betahat.W <- betahat/SE.betahat.W
p.betahat.W <- 2*(1-pt(abs(t.betahat.W),n-2))

beta.out.W <- cbind(betahat, SE.betahat.W, t.betahat.W, p.betahat.W)
colnames(beta.out.W) <- c("Estimate", "Robust Std Err", "t", "P(>|t|)")
rownames(beta.out.W) <- c("Intercept", "Budget")
round(beta.out.W, 5)

# install.packages("sandwich")
library(sandwich)
library(lmtest)

coeftest(bolly.mod1, vcov=vcovHC(bolly.mod1, type = "HC0"))
## type=:
## const => E2 = diag{s^2}  (OLS)
## HC0 => E2 = diag{e_i^2}
## HC1 => E2 = (n/(n-p))diag{e_i^2}
## HC2 => E2 = diag(e_i^2/(1-h_ii))
## HC3 => E2 = diag(e_i^2/(1-h_ii)^2)
## HC4 => E2 = diag(e_i^2/(1-h_ii)^delta_i)   with delta_i = min(4, h_ii/hbar)   with hbar = p/n
```

## Part 2e - Least Absolute Deviations Regression and Plot Data, OLS, LAD, Lowess

```{r part2e}

library(quantreg)
bolly.lad <- rq(Y1 ~ X1,0.5)
summary(bolly.lad)

## Plot Data and Regression Equations
# win.graph(height=5.5, width=7.0)
# quartz(height=5.5, width=7.0)
plot(Y1 ~ X1, pch=19, xlab="log(Budget)", ylab="log(Gross)",
  main="Bollywood Movie Revenues (Y) and Budgets (X)", cex=0.65)
abline(bolly.mod1, col="blue", lwd=2)
abline(bolly.lad, col="darkred", lwd=2)
lines(lowess(Y1, X1), col="darkgreen", lwd=2)
legend("topleft", c("Data", "OLS", "LAD", "Lowess"), pch=c(19,NA,NA,NA),
   col=c("black","blue","darkred", "darkgreen"),  lwd=c(NA,2,2,2))

```

