---
title: "Assignment3STA3100"
author: "Eric Brewster"
date: "2023-09-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Question 1}
set.seed(3100)
variables <- rexp(100, rate = 1)
M <- matrix(variables, nrow = 10, ncol = 10)
M
```


```{r Question 1a}
transposeM <- t(M)
transposeM
```


```{r Question 1b}
MtransposeM <- transposeM %*% M
MtransposeM
```


```{r Question 1c}
addIdentity <- 2.5 * M + 5.2 * diag(10)
addIdentity
```


```{r Question 1d}
inverseM <- solve(M)
inverseM
```


```{r Question 1e}
meanM = mean(M)
meanM
```


```{r Question 1f}
rowMax <- which.max(rowSums(M))
rowMax
```


```{r Question 1g}
colMeans(M>1)* 100
```


```{r Question 1h}
rownames(M) <- paste("n", 1:10, sep = "")
colnames(M) <- paste("v", 1:10, sep = "")
M
```


```{r Question 1i}
subM <- M[c("n3", "n4", "n5"), (c("v1", "v3", "v5", "v6", "v7", "v8", "v9", "v10"))]
subM
```


```{r Question 1j}
L <- diag(diag(M))
L
```


```{r Question 1k}
L <- rbind(L, rep(0, ncol(L)))
L
```


```{r Question 1l}
H <- M[ ,ncol(M)] %*% M[nrow(M), ]
H
```


```{r Question 2a}
margContTable <- margin.table(UCBAdmissions, margin = c(1, 2))
margContTable <- addmargins(margContTable)
margContTable
```


```{r Question 2b}
propTable <- prop.table(margContTable, margin = 2) * 200
propTable <- round(propTable, digits = 1)
propTable
```
For whom is the admission rate higher overall, males or females?
The admission rate is significantly higher (14.1 percent) for males than females.


```{r Question 2c}
gendDep <- ftable(prop.table(UCBAdmissions, margin = c(1, 2))) * 100
gendDep <- round(gendDep, digits = 1)
gendDep
```
How do these results compare to what you found in part (b)?
Departments A and B show the same outcome where a much higher percentage of males are admitted, as do departments C and E where far more females are rejected than males.


```{r Question 2d}
chisq.test(marginSums(UCBAdmissions, c("Dept", "Gender")))
```
What do you conclude?
Because the p-value is so low, there is a statistically significant connection between gender and department choice.

