---
title: "Assignment7STA3100"
author: "Eric Brewster"
date: "2023-10-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```


```{r Question 1}
library(tidyverse)
library(nycflights13)
worst48 <- flights %>% 
  group_by(origin, year, month, day, hour) %>%
  summarise(dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(dep_delay)) %>%
  head(48)
worst_flights <- weather %>%
  right_join(worst48, by = c("origin", "year", "month", "day", "hour")) %>%
  select(month, day, hour, temp, wind_speed, precip, dep_delay)
worst_flights %>%
  ggplot(aes(precip, dep_delay)) + geom_point() + geom_smooth()
worst_flights %>%
  ggplot(aes(wind_speed, dep_delay)) + geom_point() + geom_smooth()
```
There does not seem to be a significant correlation between delays and precipitation or wind speed

```{r Question 2}
airport_locations <- nycflights13::airports %>%
  select(faa, lat, lon)

flights %>%
  select(year:day, hour, origin, dest) %>%
  left_join(
    airport_locations,
    by = c("origin" = "faa")
  ) %>%
  left_join(
    airport_locations,
    by = c("dest" = "faa")
  )
```

It is easier to rename after the join. You do not have to save the values as variables to do a join. You merely add one line of code in the join.

```{r Question 3}
average_delay <- flights %>%
  group_by(dest) %>%
  summarise(delay = mean(arr_delay, na.rm = TRUE)) %>%
  inner_join(airports, by = c(dest = "faa"))

average_delay %>%
  ggplot(aes(lon, lat, color = delay)) +
  borders("state") +
  geom_point() +
  coord_quickmap()
```


```{r Question 5}
mdy("January 1, 2010")
ymd("2015-Mar-07")
dmy("06-Jun-2017")
mdy(c("August 19 (2015)", "July 1 (2015)"))
mdy("12/30/14")
parse_time("1705", "%H%M")
parse_time("11:15:10 PM", "%H:%M:%OS %p")
```

```{r Question 6}
calculate_age <- function(birthdate) {
  birthdate <- ymd(birthdate)
  current_date <- Sys.Date()
  age <- as.numeric(difftime(current_date, birthdate, units = "weeks")) / 52.1775
  age <- floor(age)
  return(age)
}

birthday <- "2003-03-18"
age <- calculate_age(birthday)
age
```


```{r Question 7}
boxplot(islands, main="Boxplot of Island Areas (Original Scale)")

boxplot(log(islands), main="Boxplot of Island Areas (Log Scale)")

dotchart(islands, main="Dot Chart of Island Areas")

```
Yes, a log transformation is needed to better display the dot chart.

```{r Question 8}
plot(pressure$temperature, pressure$pressure, 
     xlab="Temperature", ylab="Pressure",
     main="Scatterplot of Temperature vs. Pressure")
residuals <- with(pressure, pressure - (0.168 + 0.007 * temperature)^(20/3))
qqnorm(residuals)
qqline(residuals)
transformed_pressure <- pressure$pressure^(3/20)
plot(pressure$temperature, transformed_pressure,
     xlab="Temperature", ylab="Transformed Pressure",
     main="Scatterplot of Temperature vs. Transformed Pressure")
transformed_residuals <- transformed_pressure - (0.168 + 0.007 * pressure$temperature)^(20/3)
qqnorm(transformed_residuals)
qqline(transformed_residuals)

```

8. A linear relationship is not evident from the first graph from 8a, but is far more evident in the 8c graph.
